Mathematical description of the power system model \emph{medea}

### 2019-08-01 doc/mathematical_description.tex

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

@@

%\usepackage{gensymb}
%\usepackage{subcaption}
%\usepackage{url}
%\usepackage{breakurl}
%\def\UrlBreaks{\do\/\do-}
%
%\usepackage{multibib}
%\newcites{sec}{Bibliography} 
\begin{document}
\title{Mathematical description of the power system model \emph{medea}}
\section{Overview}
\emph{medea} is a simple, stylized and parsimonious power system model.
It simulates investment in intermittent and conventional electricity and heat generation technologies as well as in cross-border electricity transmission capacities.
At the same time, the model determines the system-cost minimizing hourly dispatch of electricity and heat generators to meet price-inelastic demand.
Model results include hourly energy generation by technology and the associated fuel use and CO2 emissions, investment in and decommissioning of conventional and renewable generators and energy storages, hourly cross-border flows of electricity and potentially required transmission capacity expansion, as well as producer and consumer surplus.

@@

% decisions covered - intermittent generation given, thermal and storage dispatch, investment in generation and storage technologies, interconnector capacity
% intermittent technologies: wind onshore \& offshore, pv, run-of-river
% dispatchable technologies: $36$ technologies for dispatchable power and heat generation, including co-generation units for the combined generation of power and heat (for industry or district heating).

%\emph{medea} numerically determines a partial-equilibrium in interconnected electricity markets as would be implemented by a social planner seeking to minimize the total system cost. 
%In particular, this implies that social planners seeking to minimize system cost of individual electricity markets might have an incentive to deviate from the overall cost minimum.
%The model assumes perfectly price inelastic demand for electricity and heat. 
%Within each country we assume sufficient transmission capacities to be in place ('copperplate'). 
%Moreover, heat demand can be met by any heat generator operating in the same country.

%It covers the hourly dispatch of thermal and hydro storage plants in seven European countries, while taking electricity generation from intermittent sources such as wind or solar energy as given. (Not true because curtailment is possible)
%In addition, the model can determine optimal investment in intermittent and dispatchable generation technologies, storage technologies, and international transmission capacities.

% It can be used to simulate short-term dispatch decisions, but also currently covers

% determines the system cost-minimizing hourly dispatch of thermal and hydro storage power plants required to meet price-inelastic (residual) demand for electricity and district heat.

% Thermal power plants and hydro storage plants are grouped in $34$ technology clusters. We differentiate the clusters by generation technologies (e.g. steam turbine, combustion turbine, combined cycle, etc.) and by fuels (uranium, lignite, hard coal, natural gas, mineral oil, biomass, water).

% Thermal power plants burn fuels to generate electricity and are constrained in their operation by installed capacities for heat and power generation.

% Heat generation is possible in heat boilers (aggregate capacity $30$ GWth, efficiency $0.9$) or in combined heat and power (CHP) plants, which must respect the limits of their feasible operation region.\footnote{The feasible operation region specifies all viable combinations of heat and power generation along with the required fuel use.}

% Heat generation and consumption are aggregated, i.e. heat generators are not serving specific district heating systems.

% Electricity generation from intermittent renewable sources (wind, solar, run-of-river) is given exogenously but can be curtailed.

% Electricity can also be stored in reservoir and pumped hydro storages.

%Generation from hydro storage plants is constrained by turbine capacity and energy contained in reservoirs of limited size.
%Reservoirs are filled by inflows or by pumping (pumped storages only).
%To better capture operational differences, we model daily, weekly and seasonal reservoir and pumped storage plants separately.
%Electricity exchange with countries outside Austria and Germany is held fixed at hourly quantities realized in 2017.
%To ensure a stable and secure operation of the electricity system, power plants must provide ancillary services such as frequency control or voltage support.
%We assume that this requires generators with an installed capacity of at least $21$ GW \footnote{This is equal to 12.5\% of peak load plus 7.5\% of installed solar and wind power capacity and is broadly in line with findings by \citet{Hirth2015} and \citet{Nicolosi2012}.} to be operational (either generating or pumping in case of pumped hydro storages) at any point in time.

%For a mathematical description see \ref{Apndx2}. The model is implemented in GAMS and was solved by Gurobi on an Intel Xeon Gold $6144$ with $264$ GB RAM.
%The model code can be found at \url{https://github.com/sebwehrle/medea}.

@@

\makecell[l]{power generating \\technologies} & $i \in I$ & \texttt{tec} &

@@

CHP technologies & $j \in I$ & \texttt{tec\_chp} & \makecell[l]{
	\texttt{lig\_stm\_chp, lig\_boa\_chp,}\\
	\texttt{coal\_sub\_chp, coal\_sc\_chp,}\\
	\texttt{coal\_usc\_chp, ng\_stm\_chp,}\\ 
	\texttt{ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp,}\\
	\texttt{ng\_cc\_lo\_chp, ng\_cc\_hi\_chp,}\\
	\texttt{ng\_mtr\_chp, ng\_boiler\_chp,}\\
	\texttt{oil\_stm\_chp, oil\_ctb\_chp,} \\ 
	\texttt{oil\_cc\_chp, bio\_chp} } \\ \hline

\makecell[l]{power to heat \\ technologies} & $h \in I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline

storage technologies & $k \in K$ & \texttt{tec\_strg} & \makecell[l]{
	\texttt{res\_day, res\_week, res\_season,} \\ 
	\texttt{psp\_day, psp\_week, psp\_season,} \\
	\texttt{battery}
	} \\ \hline

@@

energy products & $p \in P$ & \texttt{prd} & \texttt{power, heat} \\ \hline \hline
\end{tabular}

\section{Parameters} \label{parameters}
\begin{tabular}{l l l l}
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\
\hline \hline
\makecell[l]{minimal conventional \\generation} & $\mu_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(r)} & GW \\ \hline
energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(r,t,prd)} & GW \\ \hline
power plant efficiency & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & $MW / MW_{th}$\\ \hline
fuel emission intensity & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $t \ce{CO2}/MWh_{th}$\\ \hline
\makecell[l]{feasible operating \\region} & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & MWh ? \\ \hline
\makecell[l]{feasible operating \\region} & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & MWh ? \\ \hline
\makecell[l]{intermittent \\generation profile} & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(r,t,tec\_itm)} & \% \\ \hline
%\makecell[l]{exports to not \\modelled regions} & $X$ & \texttt{FLOW\_EXPORTS(r,t)} & GW \\ \hline
%\makecell[l]{imports from not \\modelled regions} & $X$ & \texttt{FLOW\_IMPORTS(r,t)} & GW \\ \hline
\makecell[l]{installed capacity \\of intermittent \\generators} & $r^{0}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(r,tec\_itm)} & GW \\ \hline
\makecell[l]{installed capacity \\of thermal generators} & $g^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(r,tec)} & GW \\ \hline
\makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(r,tec\_itm)} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\thermal generators \\(specific, annuity)} & $\kappa_{z,i}$ & \texttt{INVESTCOST\_THERMAL(r,tec)} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\storages - power \\(specific, annuity)} & $\kappa^{P}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_power')} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_energy')} & $\frac{mn EUR}{GW}$ \\ \hline
net transfer capacity & $\tau_{z,zz}$ & \texttt{NTC(r,rr)} & GW \\ \hline
\makecell[l]{count of power plants \\of same technology} & $|i_{z}|$ & \texttt{NUM(r,tec)} & . \\ \hline
fixed O\&M cost & $\kappa^o_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & EUR \\ \hline
variable O\&M cost & $\kappa^o_{i}$ & \texttt{OM\_VAR\_COST(tec)} & EUR / MWh \\ \hline
% day-ahead price & $p_{el}$ & \texttt{PRICE\_DA(t,r)} & EUR / MWh \\ \hline
\ce{CO2} price & $p^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,r)} & EUR / t CO2 \\ \hline
fuel price & $p_{t,z,f}$ & \texttt{PRICE\_FUEL(t,r,f)} & EUR / MWh \\ \hline
reservoir inflows & $\rho_{z,t,k}$ & \texttt{RESERVOIR\_INFLOWS(r,t,tec\_strg)} & MW \\ \hline
max power out & $\bar{s}^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_out')} & GW \\ \hline
max power in & $\bar{s}^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_in')} & GW \\ \hline
max energy stored & $\bar{v}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'energy\_max')} & . \\ \hline
efficiency power out & $\eta^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_out')} & . \\ \hline
efficiency power in & $\eta^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_in')} & . \\ \hline
\hline
\end{tabular}

\section{Variables} \label{variables}
\begin{tabular}{l l l l}
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\
\hline \hline
%cost variables
system cost & $C_{z}$ & \texttt{cost(r)} & EUR \\ \hline
emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(r,t,tec)} & EUR \\ \hline
fuel cost & $C^{f}_{z,t,i}$ & \texttt{cost\_fuel(r,t,tec)} & EUR \\ \hline
total o\&m cost & $C^{om}_{z,i}$ & \texttt{cost\_om(r,tec)} & EUR \\ \hline
\makecell[l]{capital cost of \\generators} & $C_{z}^{inv,i}$ & \texttt{cost\_invgen(r)} & EUR \\ \hline
\makecell[l]{capital cost of \\storages} & $C_{z}^{inv,k}$ & \texttt{cost\_invstrg(r)} & EUR \\ \hline
\makecell[l]{capital cost of \\interconnectors} & $C_{z}^{inv,ic}$ & \texttt{cost\_gridexpansion(r)} & EUR \\ \hline
% investment and decommissioning variables
\makecell[l]{added capacity of \\intermittents} & $r^{+}_{z,n}$ & \texttt{invest\_res(r,tec\_itm)} & GW \\ \hline
\makecell[l]{added capacity of \\conventionals} & $g^{+}_{z,i}$ & \texttt{invest\_thermal(r,tec)} & GW \\ \hline
\makecell[l]{added storage \\capacity (power)} & $v^{+}_{z,k}$ & \texttt{invest\_storage\_power(r,tec\_strg)} & EUR \\ \hline
\makecell[l]{added in storage \\capacity (energy)} & $s^{+}_{z,k}$ & \texttt{invest\_storage\_energy(r,tec\_strg)} & EUR \\ \hline
\makecell[l]{added transmission \\capacity} & $x^{+}_{z,zz}$ & \texttt{invest\_ntc(r,rr)} & EUR \\ \hline
\makecell[l]{decommissioned \\capacity of \\conventionals} & $g^{-}_{z,i}$ & \texttt{decommission(r,tec)} & GW \\ \hline
% dispatch variables
\makecell[l]{energy generated by \\conventionals} & $g_{z,t,i,p}$ & \texttt{q\_gen(r,t,tec,prd)} & GW \\ \hline
\makecell[l]{electricity generated \\by intermittents} & $r_{z,t,n}$ & \texttt{q\_itm(r,t,tec\_itm)} & GW \\ \hline
\makecell[l]{operating region \\weight} & $w_{z,t,i,l}$ & \texttt{cc\_weights(r,t,tec,l)} & . \\ \hline
\makecell[l]{fuel burn for energy \\generation} & $b_{z,t,i,f}$ & \texttt{q\_fueluse(r,t,tec,f)} & GW \\ \hline
energy stored in & $s^{in}_{z,t,k}$ & \texttt{q\_store\_in(r,t,tec\_strg)} & GW \\ \hline
energy stored out & $s^{out}_{z,t,k}$ & \texttt{q\_store\_out(r,t,tec\_strg)} & GW \\ \hline
\makecell[l]{storage energy \\content} & $v_{z,t,k}$ & \texttt{storage\_level(r,t,tec\_strg)} & GWh \\ \hline
\makecell[l]{electricity net \\export} & $x_{z,zz,t}$ & \texttt{flow(r,rr,t)} & GW \\ \hline
curtailed energy & $\Omega^{-}_{z,t,p}$ & \texttt{q\_curtail(r,t,prd)} & GW \\ \hline
non-served energy & $\Omega^{+}_{z,t,p}$ & \texttt{q\_nonserved(r,t,prd)} & GW \\ \hline
% other variables
\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(r)} & t \ce{CO2} \\ \hline

@@

\newpage
\section{Mathematical description} \label{mathmodel}
Objective function: total system cost, minimization
\begin{align}
\min C = \sum_{z} (C_{z})
\end{align}
System cost consist of fuel cost $C^{f}_{z,t,i}$, emission cost $C^{e}_{z,t,i}$, operation and maintenance cost $C^{om}_{z,i}$, and capital costs of investment in generation, storage and transmission equipment plus the cost of non-served load that accrues when demand is not met, i.e. when there is a black out.
\begin{align}
\begin{split}
C_{z} = \sum_{t,i} & C^{f}_{z,t,i} + \sum_{t,i} C^{e}_{z,t,i} + \sum_{i} C^{om}_{z,i} + C^{inv,i}_{z} + \\
& C^{inv,n}_{z} + C^{inv,k}_{z} + C^{inv,ic}_{z} + \sum_{t,p} C^{nse}_{z,t,p} \qquad \qquad \forall z
\end{split}
\end{align}

Our power system model uses a linear programming formulation of the economic dispatch problem for thermal units within the Austro-German bidding zone. Operation of pumped storage plants is also formulated as a linear problem.
The model's objective is to minimize total system cost, the sum of fuel, emission and operation and maintenance cost along with the cost associated with curtailment of renewable energies and loss of load.
\begin{align}
\min \left(\sum_{t,g,f}\left(\left(P_{t,f} + em_{f} P_{t,eua}+om_{g}\right) qf_{t,g,f} + qns_{t,p} M + qct_{t} N \right) \right)
\end{align}
In each hour the market has to clear, such that electricity supply from thermal and net generation from hydro storage plants plus power generation from non-dispatchable sources (wind energy, photovoltaics, and run-of-river hydro plants) equals electricity demand less net imports of electricity.
\begin{align}\label{12}
\begin{split}
D_{t,pwr} - Q_{t,nip} = \sum_{g} \left(q p_{t,g,pwr} \right) + \sum_{g\in PSP} \left(qpsp_{t,g} - ppsp_{t,g} \right) \\
+ Q_{t,we} + Q_{t,pv} + Q_{t,ror}, \forall t
\end{split}
\end{align}
In linear (economic dispatch) models, the marginals ('shadow prices') on equation (\ref{12}) can be interpreted as power prices in an energy-only market. We use these marginals to derive the pass-through from emission prices to power prices.
As we also consider co-generation of heat and power in our model, we introduce the heat balance equation (\ref{13}). Heat supply from CHP units and heat boilers must be adequate to meet district heating demand $D_{t,ht}$.
\begin{align}\label{13}
D_{t,ht} \leq \sum_{g \in CHP} \left( qp_{t,g,ht} \right) + qns_{t,p}, \forall t
\end{align}
Hourly generation of power and heat is constrained by installed capacity. 
\begin{align}
qp_{t,g,p} \leq \bar{C}_{g,p}, \forall t
\end{align}
Coproduction of heat and power in CHP-plants is governed by
\begin{align}
\sum_{l} sconv_{t,g,l} = 1, \forall g \in CHP \\
\sum_{l} sconv_{t,g,l} ORP_{g,l,p} = qp_{t,g,p}, \forall g \in CHP \\
\sum_{l} sconv_{t,g,l} ORF_{g,l,f} \leq qfuel_{t,g,f}
\end{align}
Power production by power-only generators is modelled as a linear function of plant efficiency
\begin{align}
qp_{t,g,pwr} \leq \sum_{f} \eta_{g,f,pwr} qf_{t,g,f}, \forall t, g \notin CHP  
\end{align}
Ancillary services must be provided by operating thermal generation units or active hydro storage plants (regardless of whether they are pumping or turbining).
\begin{align}
\underline{a} \leq qp_{t,g,pwr} + qpsp_{h,g} + qstor_{t,g}, \forall t
\end{align}
Operation of pumped storage plants is subject to the equations
\begin{align}
qpsp_{h,g} \eta_{g} \leq \overline{C}_{g}, \forall t, g \in PSP \\
ppsp_{t,g} \leq \overline{C}_{g}, \forall t, g \in PSP \\
qstor_{t,g} - qstor_{t-1, g} = ppsp_{h,g} \eta_{g} - qpsp_{h,g} \\
qstor_{t,g} \leq \overline{STOR}_{g}, \forall t, g \in PSP
\end{align}

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[fleqn]{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\begin{document}
\title{Mathematical description of the power system model \emph{medea}}
\section{Overview}
\emph{medea} is a simple, stylized and parsimonious power system model.
It simulates investment in intermittent and conventional electricity and heat generation technologies as well as in cross-border electricity transmission capacities.
At the same time, the model determines the system-cost minimizing hourly dispatch of electricity and heat generators to meet price-inelastic demand.
Model results include hourly energy generation by technology and the associated fuel use and CO2 emissions, investment in and decommissioning of conventional and renewable generators and energy storages, hourly cross-border flows of electricity and potentially required transmission capacity expansion, as well as producer and consumer surplus.

@@

@@

\makecell[l]{power \\generation \\technologies} & $i \in I$ & \texttt{tec} &

@@

\makecell[l]{CHP \\technologies} & $j \in J \subset I$ & \texttt{tec\_chp} & \makecell[l]{
	\texttt{lig\_stm\_chp, lig\_boa\_chp,}\\
	\texttt{coal\_sub\_chp, coal\_sc\_chp,}\\
	\texttt{coal\_usc\_chp, ng\_stm\_chp,}\\ 
	\texttt{ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp,}\\
	\texttt{ng\_cc\_lo\_chp, ng\_cc\_hi\_chp,}\\
	\texttt{ng\_mtr\_chp, ng\_boiler\_chp,}\\
	\texttt{oil\_stm\_chp, oil\_ctb\_chp,} \\ 
	\texttt{oil\_cc\_chp, bio\_chp} } \\ \hline

\makecell[l]{power to heat \\ technologies} & $h \in H \subset I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline

\makecell[l]{storage \\technologies} & $k \in K$ & \texttt{tec\_strg} & \makecell[l]{
	\texttt{res\_day, res\_week, res\_season,} \\ 
	\texttt{psp\_day, psp\_week, psp\_season,} \\
	\texttt{battery}
	} \\ \hline

@@

energy products & $m \in M=\{\text{el}, \text{ht}\}$ & \texttt{prd} & \texttt{power, heat} \\ \hline \hline
\end{tabular}

\section{Parameters} \label{parameters}
\begin{tabular}{l l l l}
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\
\hline \hline
\makecell[l]{minimal conventional \\generation} & $\sigma_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(r)} & GW \\ \hline
energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(r,t,prd)} & GW \\ \hline
power plant efficiency & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & $MW / MW_{th}$\\ \hline
fuel emission intensity & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $t \ce{CO2}/MWh_{th}$\\ \hline
\makecell[l]{feasible operating \\region} & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & MWh ? \\ \hline
\makecell[l]{feasible operating \\region} & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & MWh ? \\ \hline
\makecell[l]{intermittent \\generation profile} & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(r,t,tec\_itm)} & \% \\ \hline
%\makecell[l]{exports to not \\modelled regions} & $X$ & \texttt{FLOW\_EXPORTS(r,t)} & GW \\ \hline
%\makecell[l]{imports from not \\modelled regions} & $X$ & \texttt{FLOW\_IMPORTS(r,t)} & GW \\ \hline
\makecell[l]{installed capacity \\of intermittent \\generators} & $\bar{r}^{0}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(r,tec\_itm)} & GW \\ \hline
\makecell[l]{installed capacity \\of thermal generators} & $\bar{g}^{0}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(r,tec)} & GW \\ \hline
\makecell[l]{capital cost of \\intermittent generators \\(specific, annuity)} & $\kappa_{z,n}$ & \texttt{INVESTCOST\_ITM(r,tec\_itm)} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\thermal generators \\(specific, annuity)} & $\kappa_{z,i}$ & \texttt{INVESTCOST\_THERMAL(r,tec)} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\storages - power \\(specific, annuity)} & $\kappa^{P}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_power')} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{capital cost of \\storages - energy \\(specific, annuity)} & $\kappa^{E}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'cost\_energy')} & $\frac{mn EUR}{GW}$ \\ \hline
\makecell[l]{installed available \\transfer capacity} & $\bar{x}^{0}_{z,zz}$ & \texttt{ATC(r,rr)} & GW \\ \hline
\makecell[l]{count 100 MW slices \\of same technology} & $|i_{z}|$ & \texttt{NUM(r,tec)} & . \\ \hline
quasi-fixed O\&M cost & $o^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & EUR \\ \hline
variable O\&M cost & $o^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & EUR / MWh \\ \hline
% day-ahead price & $p_{el}$ & \texttt{PRICE\_DA(t,r)} & EUR / MWh \\ \hline
\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_EUA(t,r)} & EUR / t CO2 \\ \hline
fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,r,f)} & EUR / MWh \\ \hline
reservoir inflows & $\rho_{z,t,k}$ & \texttt{RESERVOIR\_INFLOWS(r,t,tec\_strg)} & MW \\ \hline
max power out & $\bar{s}^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_out')} & GW \\ \hline
max power in & $\bar{s}^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'power\_in')} & GW \\ \hline
max energy stored & $\bar{v}^{0}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'energy\_max')} & . \\ \hline
efficiency power out & $\eta^{out}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_out')} & . \\ \hline
efficiency power in & $\eta^{in}_{z,k}$ & \texttt{STORAGE\_PROPERTIES(r,tec\_strg,'efficiency\_in')} & . \\ \hline
value of lost load & $\mu$ & \texttt{not set yet} & EUR \\ \hline
\hline
\end{tabular}

\section{Variables} \label{variables}
\begin{tabular}{l l l l}
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\
\hline \hline
%cost variables
zonal system cost & $C_{z}$ & \texttt{cost(r)} & EUR \\ \hline
emission cost & $C^{e}_{z,t,i}$ & \texttt{cost\_emission(r,t,tec)} & EUR \\ \hline
fuel cost & $C^{f}_{z,t,i}$ & \texttt{cost\_fuel(r,t,tec)} & EUR \\ \hline
total o\&m cost & $C^{om}_{z,i}$ & \texttt{cost\_om(r,tec)} & EUR \\ \hline
\makecell[l]{capital cost of \\generators} & $C_{z}^{inv,i}$ & \texttt{cost\_invgen(r)} & EUR \\ \hline
\makecell[l]{capital cost of \\storages} & $C_{z}^{inv,k}$ & \texttt{cost\_invstrg(r)} & EUR \\ \hline
\makecell[l]{capital cost of \\interconnectors} & $C_{z}^{inv,ic}$ & \texttt{cost\_gridexpansion(r)} & EUR \\ \hline
% investment and decommissioning variables
\makecell[l]{added capacity of \\intermittents} & $\bar{r}^{+}_{z,n}$ & \texttt{invest\_res(r,tec\_itm)} & GW \\ \hline
\makecell[l]{added capacity of \\conventionals} & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(r,tec)} & GW \\ \hline
\makecell[l]{added storage \\capacity (power)} & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(r,tec\_strg)} & EUR \\ \hline
\makecell[l]{added storage \\capacity (energy)} & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(r,tec\_strg)} & EUR \\ \hline
\makecell[l]{added transmission \\capacity} & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(r,rr)} & EUR \\ \hline
\makecell[l]{decommissioned \\capacity of \\conventionals} & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(r,tec)} & GW \\ \hline
% dispatch variables
\makecell[l]{energy generated by \\conventionals} & $g_{z,t,i,p}$ & \texttt{q\_gen(r,t,tec,prd)} & GW \\ \hline
\makecell[l]{electricity generated \\by intermittents} & $r_{z,t,n}$ & \texttt{q\_itm(r,t,tec\_itm)} & GW \\ \hline
\makecell[l]{operating region \\weight} & $w_{z,t,i,l}$ & \texttt{cc\_weights(r,t,tec,l)} & . \\ \hline
\makecell[l]{fuel burn for energy \\generation} & $b_{z,t,i,f}$ & \texttt{q\_fueluse(r,t,tec,f)} & GW \\ \hline
energy stored in & $s^{in}_{z,t,k}$ & \texttt{q\_store\_in(r,t,tec\_strg)} & GW \\ \hline
energy stored out & $s^{out}_{z,t,k}$ & \texttt{q\_store\_out(r,t,tec\_strg)} & GW \\ \hline
\makecell[l]{storage energy \\content} & $v_{z,t,k}$ & \texttt{storage\_level(r,t,tec\_strg)} & GWh \\ \hline
\makecell[l]{electricity net \\export} & $x_{z,zz,t}$ & \texttt{flow(r,rr,t)} & GW \\ \hline
curtailed energy & $\Omega^{+}_{z,t}$ & \texttt{q\_curtail(r,t)} & GW \\ \hline
non-served energy & $\Omega^{-}_{z,t,p}$ & \texttt{q\_nonserved(r,t,prd)} & GW \\ \hline
% other variables
\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(r)} & t \ce{CO2} \\ \hline

@@

\newpage
\section{Mathematical description} \label{mathmodel}
\paragraph{Model objective}
\emph{medea} minimizes total system cost, i.e. the total cost of generating electricity and heat from technologies and capacities adequate to meet demand.
\begin{align}
\min C = \sum_{z} (C_{z})
\end{align}
System costs consist of fuel cost $C^{f}_{z,t,i}$, emission cost $C^{e}_{z,t,i}$, operation and maintenance cost $C^{om}_{z,i}$, capital costs of investment in generation, storage and transmission equipment, and the cost of non-served load that accrues when demand is not met, e.g. when there is a power outage.
\begin{align}
\begin{split}
C_{z} = \sum_{t,i} & C^{f}_{z,t,i} + \sum_{t,i} C^{e}_{z,t,i} + \sum_{i} C^{om}_{z,i} + C^{inv,i}_{z} + \\
& C^{inv,n}_{z} + C^{inv,k}_{z} + C^{inv,ic}_{z} + C^{nse}_{z} \qquad \qquad \forall z
\end{split}
\end{align}
Upper-case $C$ represent total cost, while $\kappa$ denotes specific, annualized capital cost of technology investment. Prices for fuels and \ce{CO2} are denoted by uppercase $P$.

The components of total cost are calculated according to equations \ref{fuel_cost} to \ref{lost_load_cost}.
\begin{align}
&C^{f}_{z,t,i}& =&\ \sum_{f} \left( P_{t,z,f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i \label{fuel_cost} \\
&C^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
&C^{om}_{z,i}& =&\ \left(|i_{z}| - \bar{g}^{-}_{z,i} + \bar{g}^{+}_{z,i}\right) o^{q}_{i} + \sum_{t,m} \left(o^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
&C^{inv,i}_{z}& =&\ \kappa_{z,i} \: \bar{g}^{+}_{z,i} \qquad \qquad &\forall z\\
&C^{inv,n}_{z}& =&\ \kappa_{z,n} \: \bar{r}^{+}_{z,n} \qquad \qquad &\forall z\\
&C^{inv,k}_{z}& =&\ \kappa^{P}_{z,k} \: \bar{s}^{+}_{z,k} + \kappa^{E}_{z,k} \: v^{+}_{z,k} \qquad \qquad &\forall z\\
&C^{inv,ic}_{z}& =&\ \kappa^{ic}_{z} \: \bar{x}^{+}_{z,zz} \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
&C^{nse}_{z}& =&\ \mu \sum_{t,m} \Omega^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
\end{align}

\paragraph{Market clearing}
In each hour, the markets for electricity and heat have to clear.
Equation \ref{market_clearing_el} ensures that the total supply from conventional and intermittent sources, and storages equals total electricity demand plus net exports, electricity stored and used for heat generation. 
Likewise, equation \ref{market_clearing_ht} clears the heat market by equating heat generation to heat demand.

\begin{align}
\sum_{i} g_{z,t,i,\text{el}} + \sum_{k} s^{out}_{z,t,k} + \sum_{n} r_{z,t,n} =
D_{z,t,\text{el}} + \sum_{i} b_{z,t,i,\text{el}} + \sum_{k} s^{in}_{z,t,k} + \sum_{zz} x_{z,zz,t} 
- \Omega^{-}_{z,t,\text{el}} + \Omega^{+}_{z,t} \qquad \forall z,t \label{market_clearing_el} \\
\sum_{i} g_{z,t,i,\text{ht}} = D_{z,t,\text{ht}} - \Omega^{-}_{z,t,\text{ht}} \qquad \forall z,t \label{market_clearing_ht}
\end{align}

In its current formulation, the model represents energy-only electricity and heat markets without capacity payments. Thus, the marginals of the market clearing equations ($\partial C / \partial D_{z,t,m}$) can be interpreted as the zonal prices for electricity and heat, respectively.

\paragraph{Thermal electricity generation}
Energy generation is constrained by available installed capacity, which can be adjusted through investment and decommissioning.
\begin{align}
g_{z,t,i,m} \leq \argmax_{l}(\psi_{i,l,m}) \left( |i_{z}| + \bar{g}^{+}_{z,i} - \bar{g}^{-}_{z,i} \right) \qquad \qquad \forall z,t,i,m
\end{align}
Generator efficiency $\eta$ determines the amount of fuel that needs to be burnt in order to generate a given amount of electricity.
\begin{align}
g_{z,t,i,\text{el}} = \sum_{f} \eta_{i,\text{el},f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \notin J
\end{align}

\paragraph{Electric heat generation}
Heat pumps and electric boilers use electricity to generate heat. The amount of electricity required to generate a given amount of heat depends on the unit's efficiency $\eta$. For heat pumps, $\eta$ is assumed to be $3$.
\begin{align}
g_{z,t,i,\text{ht}} = \sum_{f} \eta_{i,\text{ht},f} \: b_{z,t,i,f} \qquad \qquad \forall z,t,i \in H
\end{align}
